---
title: Programming with SPSS
type: docs
editor_options: 
  chunk_output_type: console
---

The databases for SPSS can be downloaded [through this link](/files/data/SPSS/MATH80667A_SPSS.zip).

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "youtube-playlist.R"))
playlist_id <- "PLUB8VZzxA8Iv41lUAoTRNoat3oCVlFY62"
slide_details <- tibble::tribble(
~title, ~youtube_id,
"Loading data and two-sample t-test", "zZI2Oq1a2Og",
"Pairwise contrasts", "AT6RsM6-EdU",
"One-way ANOVA", "7LLJWscsJrc",
"One-way ANOVA (website example)", "dWKGp1khFeE",
"Logical filters", "_So7PRQkHo4",
"Contrasts and post-hoc comparisons", "jb2nftHBgsA",
"Creating categorical variables with if-else statements", "8ba_EkcbIjw",
"Two-way ANOVA", "7Z21rYuKMp0",
"Two-way ANOVA (website example)", "lxVJAzaBHyY",
"Three-way ANOVA (website example)", "_imWUkEQVo8",
"ANCOVA (website example)", "OuNr1en9i6Y",
"Repeated measures", "VPck9Z8m9hc",
"Repeated measures (within-between design)", "1eVVFziSN_k",
"MANOVA", "FsRcZRnoCj4",
"Repeated measures using linear mixed model (Syntax)", "17-Ikn9MTpg",
"Repeated measures using linear mixed model (Menus)", "niNZcLYdxvg",
"Incomplete mixed design", "lx9WpPKXWlA",
"Nested random effects","Ww2R1dNF6DM",
"Linear mediation", "K17VesYVX9Y")
```

# Videos


```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id, example = TRUE)
```

# Code

- [One-way analysis of variance](/example/oneway.sps)
- [Two-way analysis of variance](/example/twoway.sps)
- [Three-way analysis of variance](/example/threeway.sps)
- [Analysis of covariance](/example/ancova.sps)
- [Repeated measures](/example/repeated.sps)
- [MANOVA](/example/manova.sps)
- [Repeated measures using mixed models](/example/repeatedwmixedmodels.sps)
- [Mixed models](/content/MATH80667A-10-content.sps)
- [Linear mediation](/example/linearmediation.sps)
