---
title: "Three-way analysis of variance"
linktitle: "Three-way ANOVA"
type: docs
editor_options: 
  chunk_output_type: console
execute:
  echo: true
  eval: true
  message: false
  warning: false
  cache: true
  fig-align: 'center'
  out-width: '80%'
---


```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "youtube-playlist.R"))
playlist_id <- "PLUB8VZzxA8IvjcV7Yl-OW_9KI6f_2K5HY"
slide_details <- tibble::tribble(
~title, ~youtube_id,
"Introduction", "SHhP_TfZGVM",
"Interaction plots", "I63CNxonlow",
"Marginal contrast and simple effects", "KLLBNQhD0rE",
"More contrasts and interactions", "WIoxZZ4pvlE",
"SPSS walkthrough", "_imWUkEQVo8"
)

```

# Videos

The **R** code created in the video [can be downloaded here](/example/threewayanova.R) and the [SPSS code here](/example/threeway.sps).

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id, example = TRUE)
```


# Notebook

The purpose of this little demonstration is to reproduce the results from chapter 21 and 22 of @Keppel/Wickens:2004, which were performed manually, using instead the `emmeans` package. Many of the manipulations are intricate.

We consider fictional data from @Keppel/Wickens:2004 for a study on 

> the effects of verbal feedback given during the acquisition of different types of learning material on memory tested one week later.

## Data description

There are three factors: the first is `feedback`, which indicates the type of verbal feedback received by the participant: either `none` ($a_1$) to serve as control group, `positive` ($a_2$) or `negative` ($a_3$) feedback. The second factor is related to the type of learning material, one of low-frequency words with low emotional content ($b_1$), high-frequency words with low emotional content ($b_2$) and finally high-frequency words with high emotional content ($b_3$). The third factor is the target population: either fifth graders ($c_1$) or high school seniors ($c_2$). While most of the interest is in the first two factors, the three-way ANOVA is a more efficient design if we want to study both age groups.

The response variable is the number of words remembered (`words`) after one week. The design is balanced: there are $r=5$ replications for each scenario, so $n=3 \times 3 \times 2 \times 5$ observations. If we fit the three-way model with all two ways and three way interactions, we need to estimate 19 parameters (18 means and one variance). We model each participant response assuming the measurements are independent and $Y_{ijkr} \sim \mathsf{No}(\mu_{ijk}, \sigma^2)$: this indicates that each subgroup ($a_i, b_j, c_k$) has a different (theoretical) average $\mu_{ijk}$ and a common variance $\sigma^2$. The estimates $\widehat{\mu}_{ijk}$ are simply the sample averages of each group, whereas the pooled variance estimator, $$\widehat{\sigma}^2 = \frac{1}{72}\sum_{i=1}^3 \sum_{j=1}^3 \sum_{k=1}^2 \sum_{r=1}^5 (y_{ijkr} - \widehat{\mu}_{ijk})^2,$$
is the sum of squared difference between replicate observations and their group average, divided by the residual degrees of freedom (total number of observations minus number of mean parameters).

The model can be reparametrized in terms of main effects and interaction: overall average for each factor, average of residual affect after accounting for effects of row, columns and depth and finally residual affect for the three-way. This parametrization is 
$$\begin{align*}\underset{\text{theoretical average}}{\mathsf{E}(Y_{ijkr})} &= \quad \underset{\text{global mean}}{\mu} \\ &\quad +\underset{\text{main effects}}{\alpha_i + \beta_j + \gamma_k}  \\ & \quad + \underset{\text{two-way interactions}}{(\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk}} \\ & \quad + \underset{\text{three-way interaction}}{(\alpha\beta\gamma)_{ijk}}\end{align*}$$

## Setup

What are comparisons of interest? We may be interested in the effect of feedback, for example comparing whether any form of feedback increases word retention. Thus, we could look at the marginal contrast $\mu_{1..} = \frac{1}{2}(\mu_{2..} + \mu_{3..})$, in essence treating the whole model as a one-way ANOVA but using all the data to estimate the standard deviation $\sigma$. For a balanced sample, the estimated average for example of the control group `none`, $\widehat{\mu}_{1..}$, would be the sample average of the 30 participants assigned to this experimental condition.

We can proceed similarly for the other factors. One could be interested in whether the type of learning material impacts retention (overall effect), or if there are difference between high and low emotional content for high-frequency occurrences ($b_2$ and $b_3$); amounting to ignoring all observations for the low emotion low-frequency group. 

The first step is to load the data and the packages needed for the analysis.

```{r}
# Load packages
library(dplyr)
library(ggplot2)
library(emmeans)
# Load data
data(words, package = 'hecedsm')
# Check balance
xtabs(~feedback + age + material, 
      data = words)
```

This is a 3 by 3 by 2 factorial design with $r=5$ replicates.
```{r}
model <- aov(words ~ feedback * material * age,
            data = words)
anova(model)
```

Since the data are balanced, we can use the `aov` function and look at the analysis of variance produced by `aov`.^[With unbalanced data, we would need to fit the model using `lm` and use `car::Anova` to get type 2 or 3 effects.] To get all pairwise and the triplewise interaction, we use `response ~ factor1 * factor2 * factor3` notation. Since we have replications, the full model doesn't fit the data exactly and there are residual observations to estimate the variability, but estimating reliably different variance in each of the 18 subgroup wouldn't be possible with 5 observations. We can see that the three-way interaction isn't significative, and the only two-way effect is for `material:age`.

## Interaction plot 


We can try to infer whether there is an interaction by looking at averages for each pair of variable at a time, and then at all three factors concurrently. These plots theoretically are used to demonstrate the impact of interactions, but in practice the sample estimates are noisy proxies of the true subgroup averages.

We can confirm our findings by looking at the interaction plot, showing the group average for each combination of the factors.

```{r}
words |>
  group_by(feedback, material, age) |>
  summarize(meanwords = mean(words)) |>
ggplot(mapping = aes(x = feedback,
                     y = meanwords,
                     group = material,
                     color = material)) +
  geom_line() + # connect the dots
  facet_wrap(~age) +
  labs(subtitle = "mean number of words remembered",
       y = "") +
  theme_classic() +
  theme(legend.position = "bottom")
```

These summary statistics hide the raw data: due to the discreteness of the response, which consists of counts, we can slightly perturb observations and jitter them. 


```{r}
#| label: fig-raw-words
#| fig.cap: "Individual results for an experiment on the impact of verbal feedback on retention of information."
ggplot(data = words,
         mapping = aes(x = age,
                       y = words,
                       group = material,
                       color = material)) +
  geom_point(position = position_jitterdodge(jitter.height = 0.1,
                                             dodge.width = 0.3)) +
  facet_wrap(~feedback) + # wrap by third variable
  theme_classic() + # change default theme
  theme(legend.position = "bottom") + # move caption
  labs(subtitle = "mean number of words remembered",
       y = "")
```

The high frequency/high emotion group has significantly lower retention depending on feedback, with seemingly no difference with other material type when there is no verbal feedback, but a strong decrease for positive and even stronger for negative feedback. By contrasts, there is little to no effect with high school seniors, suggesting that the experimental manipulation has limited impact for these groups.

## Main effects and contrasts

The whole purpose of Chapters 21 and 22 is to show how we can explore the three-way factorial design and extract information by treating it as a one-way ANOVA with 18 groups, a series of two-way ANOVA, etc.

We look at the estimated marginal means and start by computing the marginal mean by averaging over levels of material and age, since there did not seem to be interaction between feedback and the other variables.

As before with the `emmeans` package, we specify the variables we want to keep in `specs`, the other being 'marginalized out' by ignoring the counts. The package will print a warning since the model includes an interaction, indicating that the output is potentially misleading.

```{r}
margA <- emmeans(model, specs = "feedback")
margA
```

Since the data are balanced, this is simply the average words recalled per feedback group. We can check that this is indeed the mean of each word count by feedback by computing the summary statistics and comparing them with the output of `emmeans`:

```{r}
words |>
    group_by(feedback) |>
    summarize(mean = mean(words))
```

### Marginal contrast

We next compute a marginal contrast by comparing between no-feedback and the average of positive and negative. This amounts to treating our data as a one-way ANOVA and computing contrasts as usual.


The null hypothesis is 
$$\mathscr{H}_0: \mu_{\text{none}..} = \frac{1}{2} \left(\mu_{\text{pos}..} + \mu_{\text{neg}..}\right)$$ which can be rexpressed in terms of contrast vector as $(1, -0.5, -0.5)$ or any non-zero multiple of this solution.

```{r}
contrast(object = margA,
         method = list(interaction = c(1, -0.5, -0.5)))
```

### Simple effects

Since the two-way interaction between factors `material` and `age` is significant, we can test for differences between levels of `material` within each value of `age` (averaging over all levels of `feedback`).


To do this with `emmeans`, we first marginalize out over feedback by keeping only the two other variables in `specs`, then perform an $F$ test for each value of `age` by using `joint_tests` and specifying the analysis is conditional on `age`.

```{r}
simpleBpC <-
  emmeans(model,
          specs = c("material", "age")) |>
  joint_tests(by = "age")
simpleBpC
```

We could also look at **simple contrasts** by levels of age comparing the three material and looking at low versus high emotions. This amounts to reducing data to a two-way ANOVA in a first step, then taking each depth (age) group in turn and computing a contrast.

Using the notation $A \times B \times C$ to denote the feedback, material and age groups, this compares

$$\mathscr{H}_0: \frac{1}{2} \left(\mu_{.1k} + \mu_{.2k}\right) = \mu_{.3k}$$
for age group $k$. 

In both cases, we get two sets of tests statistics and $p$-values since there are two age groups.

```{r}
# Simple contrasts
margBC <- emmeans(model,
                  specs = c("material", "age"),
                  by = "age")
levels(margBC)
margBC |>
  contrast(method = list(contrast = c(-0.5, -0.5, 1)))
```

There is strong (unadjusted) differences between low and high emotion for fifth grades, but the effect for senior is smaller and quite uncertain.

### Interaction components

The next potential object of interest is the interaction components based on marginal means. 

@Keppel/Wickens:2004 compare the average of $b_2$ and $b_3$ between age groups, fixing the frequency and varying only the emotional content. 

The first step with `emmeans` is to specify which variables to keep, then slice by age group. This gives us a set of 3 averages per age group. Next, we set up the contrast vector to compute for each age group. The last call, to `joint_tests()`, proceeds with comparing whether the average contrast results are the same in each age group by comparing the two. There is some evidence that high emotion words has a different impact, independent of feedback. 

```{r}
emmeans(model,
        specs = c("material", "age"),
        by = "age") |>
  contrast(method = list(b2vsb3 = c(0, 1, -1))) |>
  joint_tests()
```

We could also compute differences between feedback and material. Marginalizing over age yields a total of 9 cells: we treat the resulting two-way design as a one-way design with nine groups and simply compute the contrast vector, assigning weight zero for the low-frequency and low emotion group and then looking at none (weight $1$) and the average of positive and negative feedback (weights of $-0.5$). Since we want to compare this difference between $b_2$ and $b_3$, we give similar weights for the last three groups, but instead of $(1, -0.5, -0.5)$, we flip the signs.

```{r}

emmAB <- emmeans(model, 
        specs = c("feedback", "material"))
levels(emmAB)
contr <- list("none vs feedback for b2 vs b3" =
  c(0, 0, 0, -1, 0.5, 0.5, 1, -0.5, -0.5)
)

emmAB |>
  contrast(method = contr)
```

### Three factor interaction

The most complicated type of comparison is the three factor interaction (which wasn't deemed significant). We could look at a particular combination by looking at the difference of difference. This is the same as what we just computed, but this time we will compare this difference of difference between age groups to see if it is the same.

- $A$ (`feedback`): no feedback vs feedback (`none` vs average of `pos` and `neg`), 
- $B$ (`material`): comparing low vs high emotional feedback (`b2` vs `b3`) 
- $C$: fifth graders vs seniors

```{r}
threewaycontrast <- 
  emmeans(model, 
          specs = c("feedback", "material","age")) |>
  contrast(method = list(
    contrast =c(0, 0, 0, -1, 0.5, 0.5,
                1, -0.5, -0.5, 0, 0, 0,
                1, -0.5, -0.5, -1, 0.5, 0.5)))
summary(threewaycontrast, 
        adjust = "scheffe",
        scheffe.rank = 17)
```

We can adjust for multiplicity by using Scheffé's method, but we need to set up the rank of the test: for general contrasts in a two-way slice, this would be the number of groups minus one, so 8 if we look at feedback and material, 2 if we look only at the feedback type, etc.


There is no evidence here that this setup comparison for looking at differences for feedback among high frequency groups is any different between age groups.

