---
title: "Problem set 5"
---

Complete this task individually

**Submission information**: please submit on ZoneCours

- a PDF report
- your code

## Task 1

We consider data from Experiment 2 of @Jordan:2022, who measured the confidence of participants on their ability to land successfully a plane if the pilot was incapacitated, after they were exposed to a trivially uninformative 3 minute video of a pilot landing a plane, but filmed in such a way that it was utterly useless. The authors pre-registered a comparison between experimental conditions `video` vs `no video`, and found that people watching the video answered higher for the question "How confident are you that you would be able to land the plane without dying", but there was no discernible effect for "How confident are you that you would be able to successfully land the plane as well as a pilot could", contrary to expectations. They found that the order in which the questions were asked (`order`, either pilot first, or dying first) changed the response.

The database in package `hecedsm` in **R** is labelled `JZBJG22_E2`. You can also download the [SPSS database via this link](/files/data/SPSS/JZBJG22_E2.sav).


1. A dataset is said to be *balanced* if there are the same number of people in each experimental condition. Are the data balanced over `condition` and `order`? Justify your answer.
2. Look at Figure @fig-repartition and comment on the distribution of the scores along the scale.
<!-- 3. Using Levene's test, check whether the variance in each subgroup is the same. -->
3. Perform the two-way analysis of variance for both `conf_dying` and `conf_pilot` and report the null and alternative hypotheses, the test statistic, the $P$-value and the conclusion of the test. 
4. For each response variable `conf_dying` and `conf_pilot`, perform follow-up tests. Compute simple or main effects (depending on whether the interaction is significant or not), along with 95% confidence intervals for the difference in score for `video` conditions. 

## Indications and helper code

There is no **R** script this week, but below are some instructions for fitting the different models.

### Graphics {#sec-plots}

To create a pretty plot in **R**, try the following:

```{r}
#| eval: true
#| echo: true
#| label: fig-repartition
#| fig-cap: "Half-violin (density) plots, with box plots and 95% intervals, and jittered scatterplot for confidence in landing without dying, as a function of order and condition."
# install.packages(c("ggplot2","ggdist"))
# remotes::install_github("lbelzile/hecedsm")
data(JZBJG22_E2, package = "hecedsm")
library(ggplot2)
library(ggdist)

ggplot(data = JZBJG22_E2, 
       mapping = aes(x = condition, group = order, y = conf_dying)) + 
  ggdist::stat_halfeye(adjust = .5, width = 1, 
                       .width = c(0.5, 0.95)) + # 50% (box) and 95% confidence intervals
  ggdist::stat_dots(side = "left", dotsize = .4, justification = 1.05, binwidth = .1) +
  facet_grid(~order) + 
  labs(y = "", subtitle = "confidence in landing without dying") +
  theme_classic()
```

You must repeat this for the other response variable.

In SPSS, you need to specify manually with `GGRAPH` the format. This [post](https://andrewpwheeler.com/2012/05/20/bean-plots-in-spss/) shows an example of such plots.



### Welch ANOVA {#sec-welch}

Using Levene's test, we can check whether the variance in each subgroup is the same. In this example, we find that there are significant differences in variance per subgroup, but the effect on $p$-values is negligeable, so we can safely ignore it here.

If the variance are unequal, we want to fit a model to observations in subgroup $(a_j, b_k)$ for $j=1, \ldots, n_a$ and $k=1, \ldots, n_b$, specifying that in the subgroup the average is $\mu_{jk}$ and the variance $\sigma^2_{jk}$. The resulting $F$ test from comparing overall means is typically named Welch after @Welch:1947.


In **R**, the Welch ANOVA can be obtained via `oneway.test`, but there is no possibility of follow-up for marginal means and contrasts. Rather, use the function `gls` in package `nlme` as follows.

```{r}
#| eval: false
#| echo: true
model <- nlme::gls(
  model = conf_dying ~ condition*order,
  data = JZBJG22_E2,
  weights = nlme::varIdent(form = ~ 1 | condition*order))
# Use condition*order to get a different variance for each subgroup

# ANOVA table for testing the interaction - use type 2
car::Anova(model, type = 2)
```

To obtain the analysis of variance table in **R** to an object `model` representing an ANOVA with unbalanced data, use the `car::Anova` function. Use `type=2` as argument (the default). You can pass `model` to `emmeans` as before to compute main/simple effects.

In SPSS, you must first transform the $2 \times 2$ into a one-way ANOVA with a categorical variable having $4$ levels. Then, use `oneway` and code directly the contrasts for the interaction, and either the simple effects of `condition` for each level of `order`, and the main effect of `condition`, depending on which is most suitable.
