---
title: "Problem set 11"
---

Complete this task in teams of up to three students.

**Submission information**: please submit on ZoneCours

-   a PDF report
-   your code

## Task 1

Read @Rohrer:2022 and write a summary.

## Task 2

@Liu.Rim.Min.Min:2023 postulated in their Experiment 5b that the underestimation of the appreciation of initiators relative to recipients was mediated by the degree of surprise of the recipient. Their data can be obtained from `LRMM23_S5b` in the **R** package `hecedsm`. You can also download the [SPSS database via this link](/files/data/SPSS/LRMM23_S5b.sav).

1. Fit the linear mediation model using the PROCESS macro or the `mediate` function from the `mediation` **R** package
    - use the nonparametric bootstrap with the percentile method to get confidence intervals
    - compare the coefficients with those of Figure 1 from the paper.
2. @Zhao:2010 review the typology of mediation. Identify the type of mediation (complementary, competitive or indirect only) based on coefficients.
    - complementary mediation when both direct and indirect effects are of the same sign and non-zero.
    - competitive mediation when direct and indirect effects are of opposite signs.
    - indirect-only mediation when the direct effect of $X \to Y$ is null, but the effect $X \to M \to Y$ isn't.

3. The paper does not discuss any of the model assumptions.  List the assumptions of the linear mediation model and explain how some may fail to be valid, thus casting doubt on the conclusions drawn by @Liu.Rim.Min.Min:2023.


```{r}
#| eval: false
#| echo: false
data(LRMM23_S5b, package = "hecedsm")
model1 <- lm(surprise ~ condition,
                 data = LRMM23_S5b)
model2 <- lm(appreciation ~ surprise + condition,
             data = LRMM23_S5b)
linmedmod <- mediation::mediate(
  model.m = model1, 
   model.y = model2,
   boot.ci.type = "perc",
   boot = TRUE,
   mediator = "surprise",
   treat = "condition",
   treat.value = "other",
   control.value = "self",
   sims = 5000L)
```

## Task 3

Study 4 of @Risen/Gilovich:2008 (pp. 297-299) perform a mediation analysis with a two-way ANOVA using the @Baron/Kenny:1986 methodology. 

```{r}
#| eval: false
#| echo: false
Sobel <- function(alpha, 
                  gamma, 
                  var_alpha, 
                  var_gamma){
  stat <- alpha*gamma / 
    sqrt(alpha^2*var_gamma + 
           gamma^2*var_alpha + 
           var_alpha*var_gamma)
  pval <- 2*pnorm(-abs(stat))
  return(list(stat = stat, pval = pval))
}
Sobel(alpha = 0.15, 
      gamma = -1.8, 
      var_alpha = 0.05^2, 
      var_gamma = 0.74^2)

Sobel(alpha = -0.1, 
      gamma = -2.17, 
      var_alpha = 0.04^2, 
      var_gamma = 0.67^2)
```

1. Read the description and comment on the following aspects:
    a. use of the Baron and Kenny original testing procedure^[The latter is said to be suboptimal; explain why in your words.]
    b. the plausibility of the causal model implied by the directed acyclic graph drawn in Figure 2.
2. Using the summary statistics and coefficients estimates reported, recompute Sobel's statistic^[The square of the std. error of $\widehat{\gamma}\widehat{\alpha}$ is $\widehat{\gamma}^2\mathsf{Va}(\widehat{\alpha}) + \widehat{\alpha}^2\mathsf{Va}(\widehat{\gamma}) + \mathsf{Va}(\widehat{\gamma})\mathsf{Va}(\widehat{\alpha})$, where $\mathsf{Va}(\widehat{\alpha})$ is the square of the standard error for $\widehat{\alpha}$ in the summary table of the linear regression. Some authors may be excluding the $\mathsf{Va}(\widehat{\gamma})\mathsf{Va}(\widehat{\alpha})$ term from the equation.] and the _p_-value and compare them with the values reported.^[In **R**, the $p$-value for the two-sided test can be computed via `2*pnorm(abs(stat), lower.tail = FALSE)`, where `stat` is Sobel's statistic.]
3. List the assumptions of the linear causal mediation model. Are there any check of these and, if so, do they support the claims of the authors? 
4. Can the authors successfully claim mediation considering the study uses an experimental design and randomly allocates experimental condition? Why or why not?
