---
title: "Problem set 12"
---

Complete this task individually.

**Submission information**: please submit on ZoneCours

-   a PDF report
-   your code

@Elliot:2021 attempt to replicate a study of @Flavell:1966 and study unprompted verbalization by children aged 5 to 10 in an experiment. The data containing the counts of the number of verbalization across all labs are given in the `FBC66_T1` and `MULTI21_D1` datasets. You can also download the SPSS databases via these links  ([`FBC66_T1`](/files/data/SPSS/FBC66_T1.sav) and [`MULTI21_D1`](/files/data/SPSS/MULTI21_D1.sav)).

The data from Table 1 of @Flavell:1966 is the number of children who verbalized without prompting overall in the first two experimental conditions (immediate recall and delayed recall subtasks).^[The last task, point-and-name, is not considered because it specifically instructed students to speak out loud.]

For the $\chi^2$ test to be valid, we need the expected number of counts to be at least five in each subcondition (otherwise the large sample approximation breaks down and is unreliable).

The number of students for each grade (kindergarden, second and fifth) is 20.
The rows of the table are the ordered number of instances of verbalization.

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
# Reinstall package
# remotes::install_github("lbelzile/hecedsm")
data(FBC66_T1, package = "hecedsm")
chisq.test(xtabs(count ~ grade + frequency, data = FBC66_T1))
```

1. Looking at the output, what is the conclusion of the test? 
2. The authors report a test for second versus fifth grade only. Repeat their analysis by discarding the row for kindergarden students: do the result match those reported in the paper? 
3. Usual recommendations for the chi-square test is that the expected counts in each cell be at least 5. Under the null hypothesis of **equal frequency per grade**, what is the expected number of students in each cell? *Hint: calculate the marginal proportion per column*.^[For a two-way contingency table under the null hypothesis of no interaction, the expected counts $E_{ij}$ of cell $(i,j)$ is obtained as the (total of row $i$) times (total of column $j$), divided by the overall count. If the expected number is less than 5, the asymptotic $\chi^2$ approximation is dubious.  We could resort to permutation and simulation to assess the accuracy of the $\chi^2$ benchmark. The row sum for the counts (per grade) is fixed to 20, but the columns are allowed to vary.]

```{r}
#| echo: false
#| eval: false
coin::independence_test(frequency ~ grade, 
                        weights = ~count, 
                        data = FBC66_T1, 
                        teststat = "quadratic",
                        distribution = "approximate")
```

4. The replication @Elliot:2021 consider children of different ages corresponding roughly to @Flavell:1966, but with an additional year. Repeat the test and the pairwise comparisons (*Hint: in **R**, use `xtabs` to cross-classify counts of factor*).
5. Why is the $p$-value so small in their replication?
