---
title: "Introduction"
---

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- ""

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "", "Introduction", "1", TRUE,
  "", "Motivation", "motivation", FALSE,
  "", "Review", "review", FALSE,
  "", "Key concepts", "keyconceptsexperimental", FALSE
)
```

## Content

- Syllabus and learning objectives
- Review
    - Population and samples
    - Observational versus experimental studies
    - Sampling schemes: random, stratified and cluster sampling
- Introduction to experimental designs
- Terminology of experimental design
- Requirements for a good experiment


## Learning objectives

- Learning the terminology associated to experiments.
- Assessing the generalizability of a study based on the consideration of the sample characteristics, sampling scheme and population.
- Distinguishing between observational and experimental studies.
- Understanding the rationale behind the requirements for good experimental studies.

## Preliminaries

- Familiarize yourself with the [syllabus](/syllabus), [content](/content/), [examples](/example/), and [evaluations](/evaluations/) pages for this class.
- {{< fa book >}} Read Chapter 1 (Intro to Data) of [OpenIntro Statistics](https://www.openintro.org/book/os/) and the accompanying {{< fa brands youtube >}} [videos](https://www.youtube.com/playlist?list=PLkIselvEzpM6pZ76FD3NoCvvgkj_p-dE8)


## Readings

:::{.callout-warning}
These readings should be completed before class, to ensure timely understanding and let us discuss the concepts together through various examples and case studies --- the strict minimum being the course notes. If you feel a section is redundant and overlaps with the latter, feel free to skim through the rest of the material.
:::

- {{< fa book >}} Chapter 1 of the [Course notes](https://lbelzile.github.io/math80667a/01-introduction.html)
- {{< fa book >}} Chapter 2 of [*Introduction to Modern Statistics*](https://openintro-ims.netlify.app/data-design.html)

Feel free to inform me in the weekly check-in of your take on these references.

## Complementary readings

:::{.callout-warning}
Complementary readings are additional sources of information that are not required readings, but may be useful substitutes. Sometimes, they go beyond the scope of what we cover (in more details).
:::


- {{< fa book >}} Chapter 1 (*Preliminaries*) in *Planning of experiments*  [@Cox:1958]. Out of print, but addresses the basic concepts using a variety of examples (mostly from agricultural field trials), and particularly well written.
- {{< fa book >}} Chapter 1 and Sections 2.1-2.2 in *Design and Analysis of Experiments*  [@Dean:2017]

## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/01-slides")
# slide_tabs(slide_details, "/slides/01-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/01-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::

## Case study

We will discuss the summary of @Mask:2022 ["The Impact of Mask Distribution and Promotion on Mask Uptake and COVID-19 in Bangladesh"](https://www.povertyactionlab.org/evaluation/impact-mask-distribution-and-promotion-mask-uptake-and-covid-19-bangladesh) ([pdf](https://web.archive.org/web/20221126024634/https://www.poverty-action.org/printpdf/47486)) in class.

During the activity, you will be asked to identify in teams the following:

- the objective of the study
- the target population (which findings generalize?)
- the sampling scheme
- the observational and experimental units
- the treatments
- the outcome variable(s)

