---
title: "Hypothesis testing"
---

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- ""

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "", "Introduction", "1", TRUE,
  "", "Variability", "signal-vs-noise", FALSE,
  "", "Hypothesis testing", "hypothesis-tests", FALSE
)
```


## Content

- Sampling variability
- Hypothesis testing
- Pairwise comparisons


## Learning objectives

At the end of the session, students should be capable of

- understanding the mechanics behind generic hypothesis tests
- interpreting the output of generic tests
- correctly reporting the output of a testing procedure


## Readings

- {{< fa book >}}  Chapter 5 (*Foundations for inference*) of [Matthew Crump's course notes](https://www.crumplab.com/statistics/foundations-for-inference.html). These notes are non-technical, but do a good job at explaining the notion of sampling variability and chance. If you find them too basic, skip directly to the next item.
- {{< fa book >}} Chapter 2 of the [Course notes](https://lbelzile.github.io/math80667a/02-hypothesis_testing.html)
- {{< fa file >}}  [The permutation test by Jared Wilson](https://www.jwilber.me/permutationtest/)

## Complementary readings

- {{< fa book >}}  Chapter 2 of @Maxwell.Delaney.Kelley:2017
- {{< fa book >}}  Chapter 3 of @Keppel/Wickens:2004.


## Slides

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/02-slides")
# slide_tabs(slide_details, "/slides/02-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/02-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::

## Case study

We will look at the way authors report the conclusion of their statistical tests with

- @Rosen:1974
- @Brucks.Levav:2022
- @Liu.Rim.Min.Min:2023, Experiment 1

## Code

The following code reproduces the applications in the course slides and the case study

- [R script](/content/MATH80667A-02-content.R)
- [SPSS scripts](/content/MATH80667A-02-content-SPSS.zip)
- [Python script](/content/MATH80667A-02-content.py) for the case study


## References
