---
title: "Analysis of variance"
---

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- ""

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "", "Introduction", "1", TRUE,
  "", "Recap", "recap", FALSE,
  "", "One way ANOVA", "f-test", FALSE,
  "", "Parametrizations and interpretation", "parametrization", FALSE,
  "", "Planned comparisons and post-hoc tests", "planned-comparisons", FALSE
)
```


## Content

- ANOVA and _F_-test statistic
- Model assumptions


## Learning objectives

At the end of the session, students should be capable of

- carry-out a one-way analysis of variance test
- check model assumptions using hypothesis tests and graphics.


## Readings

- {{< fa book >}} Chapter 3 of the [Course notes](https://lbelzile.github.io/math80667a/03-completely_randomized_trials.html)


## Complementary readings

- {{< fa book >}}  Chapter 3 of @Maxwell.Delaney.Kelley:2017 (up to page 113).
- {{< fa book >}}  [Chapter 2](https://stat.ethz.ch/~meier/teaching/anova/completely-randomized-designs.html) of @Meier:2022
- {{< fa book >}}  Chapter 3 and 7 of @Keppel/Wickens:2004.


## Slides

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/03-slides")
# slide_tabs(slide_details, "/slides/03-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/03-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::

## Code

- [R script](/content/MATH80667A-03-content.R)
- [SPSS script](/content/MATH80667A-03-content.sps)


## References
