---
title: "Contrasts and multiple testing"
date: "2022-09-23"
---

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- ""

```


## Content

- Contrasts
- Multiple testing


## Learning objectives

At the end of the session, students should be capable of

- specifying and calculating custom contrasts in factorial designs
- determining the number of tests in a family that need to be corrected for
- understanding how to correct _p_-values to account for multiple testing
- listing multiplicity testing methods suitable depending on context


## Readings

- {{< fa book >}}  [Chapter ](https://stat.ethz.ch/~meier/teaching/anova/contrasts-and-multiple-testing.html) of @Meier:2022
- {{< fa book-open-reader >}} [One-way ANOVA example](/example/onewayanova.html)


## Complementary readings


- {{< fa book >}}  Chapters 4 and 6 of @Keppel/Wickens:2004.


## Slides

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/04-slides")
# slide_tabs(slide_details, "/slides/04-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/04-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::



## References
