---
title: "Complete factorial designs"
---

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- ""

```


## Content

- Benefits of factorial designs
- Interactions
- Simple, marginal and interaction contrasts


## Learning objectives

At the end of the session, students should be capable of

- listing the benefits of factorial designs
- interpreting an interaction plot and being aware of it's limitation
- interpreting model effects in sum-to-zero parametrization
- reporting results from a two-way analysis of variance
- determining whether or not simple or main effects are most suitable


## Readings

- {{< fa book >}}  [Chapter 4](https://stat.ethz.ch/~meier/teaching/anova/factorial-treatment-structure.html) of @Meier:2022
-  {{< fa book >}}  [Chapter 5](https://lbelzile.github.io/math80667a/05-interactions_multiway.html) of the course notes


## Complementary readings

- {{< fa book >}}  Chapter 7 of @Maxwell.Delaney.Kelley:2017.
- {{< fa book >}}  Chapters 10-13 of @Keppel/Wickens:2004.


## Slides

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/05-slides")
# slide_tabs(slide_details, "/slides/04-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/05-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::

## Code


- [R script](/content/MATH80667A-05-content.R)
- [SPSS script](/content/MATH80667A-05-content.sps)


## References
