---
title: "Multiway factorial designs"
---




## Content

- Unbalanced designs
- Multiway factorial designs


## Learning objectives

At the end of the session, students should be capable of

- listing the benefits of balanced designs
- correctly interpreting analysis of variance table
- specifying and calculating custom contrasts in multiway factorial designs
- correcting for multiple testing

## Readings

-  {{< fa book >}}  [Chapter 5](https://lbelzile.github.io/math80667a/05-interactions_multiway.html) of the course notes
- {{< fa book-open-reader >}} [Two-way ANOVA example](/example/twowayanova.html)
- {{< fa book-open-reader >}} [Three-way ANOVA example](/example/threewayanova.html)

## Complementary readings

- {{< fa book >}}  Chapter 8 of @Maxwell.Delaney.Kelley:2017.
- {{< fa book >}}  Chapters 21-22 of @Keppel/Wickens:2004.

## Videos


<div class="ratio ratio-16x9">
<iframe src="https://www.youtube.com/embed/Rss0bSmONz0" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>
</iframe>
</div>


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/06-slides")
# slide_tabs(slide_details, "/slides/06-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/06-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::


## Code


- [R script](/content/MATH80667A-06-content.R)
- [SPSS script](/content/MATH80667A-06-content.sps)


