---
title: "Multiway and multivariate factorial designs"
---

## Content

- ANOVA with two or more factors
- ANOVA with both within-subject and between-subject factors
- Multivariate analysis of variance





## Learning objectives

At the end of the session, students should be capable of

- analyzing models with both within-subject and between-subject factors
- specifying and calculating custom contrasts in multiway factorial designs
- correctly interpreting an analysis of variance table
- testing model assumptions in multivariate model

## Readings

- {{< fa book >}} [Chapter 7](https://lbelzile.github.io/math80667a/multiway.html) of the course notes
- {{< fa book >}} [Chapter 8](https://lbelzile.github.io/math80667a/manova.html) of the course notes
- {{< fa book-open-reader >}} [Three-way ANOVA example](/example/threewayanova.html)
- {{< fa book-open-reader >}} [MANOVA example](/example/manova.html)


## Complementary readings

- {{< fa book >}}  Chapters 8 and 14 of @Maxwell.Delaney.Kelley:2017.
- {{< fa book >}}  Chapters 20--23 of @Maxwell.Delaney.Kelley:2017.





## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/07-slides")
# slide_tabs(slide_details, "/slides/07-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/07-slides.html",
  ratio = "16:9"
)
```

## Code

- [R script](/content/MATH80667A-07-content.R)
- [SPSS script](/content/MATH80667A-07-content.sps)

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
