---
title: "Designs to reduce the error"
---

## Content

- Blocking
- Analysis of covariance


## Learning objectives

At the end of the session, students should be capable of

- explain the benefits of blocked design and ANCOVA over completely randomized designs
- using control covariates or blocking factors
- testing equality of slopes

## Readings

- {{< fa book >}} [Chapter 6](https://lbelzile.github.io/math80667a/06-blocking_ancova.html) of the course notes
- {{< fa book >}} Chapter 5 ([Complete Block Designs](https://stat.ethz.ch/~meier/teaching/anova/block-designs.html)) of @Meier:2022 (5.1-5.2)

## Complementary readings

- {{< fa book >}}  Chapter 9 of @Maxwell.Delaney.Kelley:2017.
- {{< fa book >}}  Chapter 4 of @Cox:1958.

## Videos


<div class="ratio ratio-16x9">
<iframe src="https://www.youtube.com/embed/gTLp0pplRXk" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>
</iframe>
</div>


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/07-slides")
# slide_tabs(slide_details, "/slides/06-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/07-slides.html",
  ratio = "16:9"
)
```

## Code

- [R script](/content/MATH80667A-07-content.R)

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
