---
title: "Effect size and power"
date: "2022-10-28"
---

## Content

- Measures of effect size
- Power calculations
- Interplay between sample size, effect and power


## Learning objectives

At the end of the session, students should be capable of

- correctly report effect size for common statistics in analysis of variance models
- deduce the sample size necessary to replicate a study at a given power
- explain the interplay between sample size, power and effect size.

## Readings

-  {{< fa book >}}  [Chapter 7](https://lbelzile.github.io/math80667a/effect-sizes-and-power.html) of the course notes
- {{< fa book-open-reader >}} [Effect size and power example](/example/effectsizepower.html)

## Complementary readings

- {{< fa file >}} @Lakens:2013
- {{< fa file >}} @Steiger:2004
- {{< fa file >}} @Kelley/Preacher:2012

## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/08-slides")
# slide_tabs(slide_details, "/slides/06-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/08-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
