---
title: "Analysis of covariance and moderation"
---

## Content

- Analysis of covariance and linear regression models
- Moderation analysis

## Learning objectives

At the end of the session, students should be capable of

- explain the benefits ANCOVA over completely randomized designs
- using control covariates
- testing equality of slopes
- testing for moderating variables (interactions)

## Readings

-  {{< fa book >}}  [Chapter 9](https://lbelzile.github.io/math80667a/ancova.html) of the course notes


## Complementary readings



- {{< fa book >}}  Chapter 6 of @Maxwell.Delaney.Kelley:2017.
- <i class="fas fa-book"></i> Chapter 14--15 of @Keppel/Wickens:2004


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/08-slides")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/08-slides.html",
  ratio = "16:9"
)
```

## Videos


<div class="ratio ratio-16x9">
<iframe src="https://www.youtube.com/embed/gTLp0pplRXk" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>
</iframe>
</div>

## Code

- [R script](/content/MATH80667A-08-content.R)
- [SPSS script](/content/MATH80667A-08-content.sps)


:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
