---
title: "Repeated measures and multivariate ANOVA"
date: "2022-11-04"
---

## Content

- Analysis of within-subject designs
- Repeated measures and test of sphericity
- Multivariate analysis of variance


## Learning objectives

At the end of the session, students should be capable of

- adequately modelling repeated measures
- correctly interpreting tests of sphericity using software
- explaining the benefits and disadvantages of repeated-measure designs
- testing model assumptions in multivariate model

## Readings

-  {{< fa book >}}  [Chapter 9](https://lbelzile.github.io/math80667a/repeated-measures-and-multivariate-models.html) of the course notes


## Complementary readings

<!--
- <i class="fas fa-book"></i>[Chapter 6 Random and Mixed-Effects Models](https://stat.ethz.ch/~meier/teaching/anova/random-and-mixed-effects-models.html) of Lukas Meier's ANOVA and Mixed Models: A Short Intro Using R] 
- [Chapter 11 of @Oehlert:2010 ](http://users.stat.umn.edu/~gary/book/fcdae.pdf)

-->

- <i class="fas fa-book"></i> Chapter 16 to 18 of @Keppel/Wickens:2004


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/09-slides")
# slide_tabs(slide_details, "/slides/06-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/09-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
