---
title: "Repeated measures and multivariate ANOVA"
---

## Content

- Analysis of within-subject designs
- Repeated measures and test of sphericity
- Multivariate analysis of variance


## Learning objectives

At the end of the session, students should be capable of

- adequately modelling repeated measures
- correctly interpreting tests of sphericity using software
- explaining the benefits and disadvantages of repeated-measure designs
- testing model assumptions in multivariate model

## Readings

-  {{< fa book >}}  [Chapter 9](https://lbelzile.github.io/math80667a/09-repeated.html) of the course notes


## Complementary readings


- {{< fa book >}} @WebPower, pages 143--148 [of the web manual](https://webpower.psychstat.org/wiki/_media/grant/webpower_manual_book.pdf) 
- {{< fa book >}}  Chapters 11--13 of @Maxwell.Delaney.Kelley:2017.
- <i class="fas fa-book"></i> [Chapter 5.2 ](https://opentext.wsu.edu/carriecuttler/chapter/experimental-design/) of @Price:2017
- <i class="fas fa-book"></i> Chapter 16 to 18 of @Keppel/Wickens:2004


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/09-slides")
# slide_tabs(slide_details, "/slides/06-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/09-slides.html",
  ratio = "16:9"
)
```

## Code

- [R script](/content/MATH80667A-09-content.R)
- [SPSS script](/content/MATH80667A-09-content.sps)

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
