---
title: "Effect size and power"
---

## Content

- Measures of effect size
- Power calculations
- Interplay between sample size, effect and power


## Learning objectives

At the end of the session, students should be capable of

- correctly report effect size for common statistics in analysis of variance models
- deduce the sample size necessary to replicate a study at a given power
- explain the interplay between sample size, power and effect size.

## Readings

- {{< fa book >}}  [Chapter 10](https://lbelzile.github.io/math80667a/power_effect.html) of the course notes
- {{< fa book >}}  [Chapter 11](https://lbelzile.github.io/math80667a/reproducibility_crisis.html) of the course notes
- {{< fa book-open-reader >}} [Effect size and power example](/example/effectsizepower.html)
- {{< fa book >}} @WebPower, pages 143--148 [of the web manual](https://webpower.psychstat.org/wiki/_media/grant/webpower_manual_book.pdf)

## Complementary readings

- {{< fa book >}} Chapters 3 (section *Power of the $F$ Test*) and 4 (section *Measures of Effects*) of @Maxwell.Delaney.Kelley:2017.
- {{< fa file >}} @Lakens:2013
- {{< fa file >}} @Steiger:2004
- {{< fa file >}} @Kelley/Preacher:2012

## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/09-slides")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/09-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::

## Exercise

- [Power calculations](/example/power.html)

## Code


- [R script](/content/MATH80667A-09-content.R)
- [Exercise solution](/example/power.R)


