---
title: "Intro to mixed models"
date: "2022-11-11"
---

## Content

- Introduction to mixed models

## Learning objectives

At the end of the session, students should be capable of

- determining the structure of an experimental design with random and fixed effects
- correctly determining whether effects are crossed or nested
- setting up a linear mixed model



## Readings

- <i class="fas fa-book"></i>[Chapter 6 Random and Mixed-Effects Models](https://stat.ethz.ch/~meier/teaching/anova/random-and-mixed-effects-models.html) of Lukas Meier's ANOVA and Mixed Models: A Short Intro Using R] 
-  {{< fa book >}}  [Chapter 10](https://lbelzile.github.io/math80667a/introduction-to-mixed-models.html) of the course notes



## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/10-slides")
# slide_tabs(slide_details, "/slides/10-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/10-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
