---
title: "Introduction to causal inference"
---

## Content

- Introduction to causal inference


## Learning objectives

At the end of the session, students should be capable of

- understanding the importance of listing potential confounders
- determining which variables to control for (confounders vs colliders)
- drawing a directed acyclic graph describing the interrelation between variables
- explaining the differences between experimental and observational studies for studying mediation


## Readings

- <i class="fas fa-book"></i>Chapter 2 of @Vanderweele:2015

Andrew Heiss' course notes on directed acyclic graphs (DAG) and types of association. 

- Causal models ([video](https://www.youtube.com/watch?v=kJQuH3nssfI&list=PLS6tnpTr39sFVrHdTifzlyS9dsJjhEhsF) and [slides](https://evalf21.classes.andrewheiss.com/slides/04-slides.html#causal-models))
- Paths, doors, and adjustment ([video](https://www.youtube.com/watch?v=_qs_1B4ySWY) and [slides](https://evalf21.classes.andrewheiss.com/slides/04-slides.html#58))

The structural equation modelling (SEM) approach to mediation

- Paper popularizing linear mediation [@Baron/Kenny:1986]
- Limitations of the linear mediation model approach [@Bullock.Green.Ha:2010]

The causal inference approach

- @Pearl:2014
- @Imai:2010

## Complementary readings

- @Pearl:2016, Chapter 3
- @Rohrer:2018


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/11-slides")
# slide_tabs(slide_details, "/slides/11-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/11-slides.html",
  ratio = "16:9"
)
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
