---
title: "Mediation and moderation"
---

## Content

- Linear mediation model
- Moderation


## Learning objectives

At the end of the session, students should be capable of

- Using the linear mediation model
- discussing limitations of the linear mediation model
- testing for moderating variables (interactions)


## Readings

- <i class="fas fa-book"></i>Chapter 11 of the [course notes](https://lbelzile.github.io/math80667a/11-causal-inference.html)
- <i class="fas fa-book"></i>Chapter 11 of @Vanderweele:2015

- Linear mediation model [@Baron/Kenny:1986]
- Limitations of the linear mediation model approach [@Bullock.Green.Ha:2010]
- Model assumptions [@Pearl:2014;@Imai:2010]


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/12-slides")
# slide_tabs(slide_details, "/slides/12-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/12-slides.html",
  ratio = "16:9"
)
```


## Code

- [R script](/content/MATH80667A-12-content.R)
- [SPSS script](/content/MATH80667A-12-content.sps)


:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
