---
title: "Nonparametric tests and count data"
date: "2022-12-02"
---

## Content

- Nonparametric tests
- Models and tests for count data


## Learning objectives

At the end of the session, students should be capable of

- Understand the benefits and drawback of rank-based tests
- Perform tests for count data (Fisher exact test, $\chi^2$ tests) and correctly report their output with effect sizes



## Readings

- {{< fa book >}}[Chapter 12 *Categorical data analysis*](https://learningstatisticswithr.com/book/chisquare.html) of @Navarro:2019
-  [Chapter 12](https://lbelzile.github.io/math80667a/nonparametric-tests.html) of the course notes


## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

<!--

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/13-slides")
# slide_tabs(slide_details, "/slides/11-slides.html")
```

```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/13-slides.html",
  ratio = "16:9"
)
```

-->


:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
