---
title: "Nonparametric tests and count data"
---

## Content

- Models and tests for count data
- Nonparametric tests


## Learning objectives

At the end of the session, students should be capable of

- Understanding the benefits and drawback of rank-based tests.
- Performing tests for count data and correctly report their output.



## Readings

- {{< fa book >}}[Chapter 12 *Categorical data analysis*](https://learningstatisticswithr.com/book/chisquare.html) of @Navarro:2019
-  Chapters [12](https://lbelzile.github.io/math80667a/12-nonparametric.html) and [13](https://lbelzile.github.io/math80667a/13-count.html) of the course notes



## Slides

```{r}
#| label: slides-videos
#| echo: false
#| include: false
source(here::here("R", "slide-things.R"))
```

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/13-slides")
```


```{r show-slides-xaringan, echo=FALSE}
xaringanExtra::embed_xaringan(
  url = "/slides/13-slides.html",
  ratio = "16:9"
)
```


## Code

- [R script](/content/MATH80667A-13-content.R)
- [SPSS script](/content/MATH80667A-13-content.sps)


:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of slide-specific commands.
:::
