---
title: "MATH 80667A - Week 10"
author: "Léo Belzile"
format: html
eval: true
echo: true
message: false
warning: false
code-tools:
      source: true
      toggle: false
      caption: "Download Quarto file"
---


```{r}
#| echo: false
options(digits = 3, knitr.kable.NA = '')
```

# Blocking factor

```{r}
#| label: tbl-anova-blocking
#| tbl-cap: "Analysis of variance table"
#| tbl-subcaption: ["with blocking factor","without the blocking factor"]
#| layout-ncol: 2
library(emmeans)
library(hecedsm)
library(ggplot2)
library(afex)
library(lmerTest)
# Force sum-to-zero parametrization for unordered factors
options(contrasts = c("contr.sum", "contr.poly"))
## Blocking factor
# Note that this is fundamentally repeated measures
url <- "https://edsm.rbind.io/files/data/resting_metabolic_rate.txt"
# transform integers to factors (categorical)
resting <- read.table(url, header = TRUE) |>
  dplyr::mutate(
    subject = factor(subject), #blocking factor
    protocol = factor(protocol), #experimental factor
    rate = rate/1000)

# Fit model with blocking factor
model_block <- aov(rate ~ subject + protocol, data = resting)
# One-way ANOVA (no blocking)
model_raw <- aov(rate ~ protocol, data = resting)

# ANOVA tables with and without blocking factor
anova(model_block) |>
  knitr::kable()
anova(model_raw) |>
  knitr::kable()
```



```{r}
# Use ANOVA table to compute etasquared effect size
(etasq <- 0.0359/(23.1175 + 1.2355 + 0.0359))
efs <- effectsize::eta_squared(
  model_block, 
  partial = TRUE,
  generalized = "subject")
```

```{r}
#| label: fig-interaction-metabolic
#| fig-cap: "Interaction plot"
# Interaction plot
ggplot(data = resting,
       aes(x = subject,
           y = rate,
           group = protocol,
           color = protocol)) +
  geom_line(linewidth = 1.5) +
  labs(subtitle = "mean resting metabolic rate",
       y = "",
       x = "subject identifier") +
  scale_color_grey()+
  theme_classic() +
  theme(legend.position = "bottom")
```

# Mixed model

## Hosano et al. (2022)

The purpose of this is to show the equivalence of repeated measures and mixed models for tests of interest
in balanced designs with a single repeated measurement per individual, here captured through individual-specific random effects.

We consider three different models for `imscore` as a function of `waiting` (between-subject factor) and `ratingtype` (within-subject factor). Each individual `id` gives two ratings. The `id` are nested withing `waiting`, but crossed with `ratingtype`.

- a mixed ANOVA (within-between) full-factorial design (i.e., including the interaction term),
- a MANOVA with the two measurements of `imscore` by `ratingtype` as response,
- a linear mixed model with a random effect for `id` and both `waiting`, `ratingtype` and their interaction.

```{r}
#| layout-ncol: 3
data(HOSM22_E3, package = 'hecedsm')
# Pivot from long to wide for MANOVA
HOSM22_E3w <- HOSM22_E3 |> tidyr::pivot_wider(
  names_from = ratingtype,
  values_from = imscore
)

xtabs(~ratingtype + waiting, HOSM22_E3) |>
  knitr::kable() # crossed factors
head(xtabs(~id + waiting, HOSM22_E3), n = 5) |>
  knitr::kable(row.names = TRUE) # id nested in waiting
head(xtabs(~id + ratingtype, HOSM22_E3), n = 5) |>
  knitr::kable(row.names = TRUE)
```


```{r}
mod <- aov_ez(id = "id",
              dv = "imscore",
              between = "waiting",
              within = "ratingtype",
              data = HOSM22_E3)

# Obtain MANOVA table
MANOVA_tab <- mod$Anova # same as below
```

```{r}
#| label: tbl-MANOVA
#| tbl-cap: "Type III Repeated Measures MANOVA Tests: Pillai test statistic"
# Fit instead the model with MANOVA using multivariate linear regression
manova_mod <- lm(cbind(prediction, experience) ~ waiting,
                 data = HOSM22_E3w)
# For repeated measures, we need to reconstruct the missing factor(s)
# corresponding to the repeated measures via a data frame (idata)
# that contains the factor levels and the variable name
# and idesign that includes the additional factors to our models
car::Anova(manova_mod,
           idata = data.frame(
             ratingtype = factor(c("prediction","experience"))),
           idesign =~ratingtype, type = 3) 
```

```{r}
# Fitting the same 2x2 model (with interaction), including a random intercept per subject
library(lmerTest)
mixmod <- lmer(
  imscore ~ waiting*ratingtype + (1 | id),
  data = HOSM22_E3)
# We get the same table if we set type III
anova(mixmod, type = 2)
```

## Curley et al.

This is an incomplete block design: there are two (counterbalanced) combinations of `anchor`, `vignette`, and `verdictsyst`, but participants see 2 out of 8 combinations. Hence, no interaction between these and ID is possible.

```{r}
data(C22, package = "hecedsm")
head(C22)
options(contrasts = c("contr.sum", "contr.poly"))
# balanced!
xtabs(~ anchor + vignette + verdictsyst, data = C22)

xtabs(~  interaction(anchor, vignette, verdictsyst) + id, data = C22)[,1:5]
model <- lmer(
  guilt ~ anchor*vignette*verdictsyst + pjaq + (1|id),
  data = C22)
# pjaq is a covariate (so used to reduce error, plus the slope is of interest on it's own
# Cannot have interaction pjaq * id, because we get a single pjaq score per person

# No ambiguity for sum of square decomposition
anova(model)
# No three-way interaction
# A two-way interaction between vignette:verdictsyst
# Computing differences between anchors
emmeans(model, specs = "anchor") |> pairs()
# Computing differences in verdict separately for each vignette
emmeans(model, specs = "verdictsyst", by = "vignette") |> pairs()
```

## Chocolate rating

```{r}
book.url <- "http://stat.ethz.ch/~meier/teaching/book-anova"
chocolate <- read.table(file.path(book.url, "data/chocolate.dat"),
                        header = TRUE)
chocolate[,"rater"]      <- factor(chocolate[,"rater"])
chocolate[,"background"] <- factor(chocolate[,"background"])
str(chocolate)
# Fit the model (note that rater recycles the id 1:10, so we need to be careful here!
chocolate <- chocolate |> dplyr::mutate(id = factor(paste(background, rater)))
# This model is correct
model <- lmer(y ~ background*choc +
                (1 | rater:background) + (1 | rater:choc:background),
              data = chocolate)
# This is fine too (because of the distinct IDs)
model <- lmer(y ~ background*choc +
                (1 | id) + (1 | choc:id),
              data = chocolate)
# This is WRONG (compare degrees of freedom)
# model <- lmer(y ~ background*choc +
#                 (1 | rater) + (1 | choc:rater),
#               data = chocolate)
# Data are again balanced
anova(model)
# There is a no evidence of interaction
# The model output includes variance coefficients
summary(model)
# Look at best chocolate type overall
(emm <- emmeans(model, specs = "choc"))
emm |> contrast("pairwise")
# C has the highest rating, but indistinguishable from A
# B is worst

# Compare variability - extract variance from model
(vars <- c(unlist(VarCorr(model)), sigma(model)^2))
# Compare these with the output of summary
# Correlation between same chocolate/rater
sum(vars[1:2])/sum(vars)
# Correlation between measurements from same rater, different chocolates
vars[2]/sum(vars)
```
